BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-09 conclusions.tex
First Draft Corrected WOUT JoeBox
First Batch of fixes (Emilio's)
Fixes Done

\chapter{Conclusions}

In order to deal with the astounding amount of malware instances surfacing every day, there has been a shift towards the use of dynamic malware analysis techniques over static ones. A popular dynamic analysis approach is function call monitoring, as it enables the analyst to grasp the behavior of the malware sample under analysis, despite the presence of code obfuscations. Such technique is normally implemented via API hooking, that is, the interception of functions exported by dynamically linked libraries (DLL). However, API hooking has to face the growing trend of evasive malware strains, which conceal their damaging behavior when under the magnifying glass of an analyst. Furthermore, most of the current API hooking techniques presented in literature do not provide any countermeasures against evasive malware.

Motivated by this realization, in this thesis we have developed BlueTracer, a robust library and system call tracer for Windows applications specialized in evasive malware, which was implemented adopting the Intel Pin dynamic binary instrumentation framework and is capable of remaining hidden thanks to its integration with the BluePill stealthy execution toolkit. During the tool's development, a notable implementation challenge was posed by the heterogeneity of Windows libraries employed by malware and by the lack of appropriate documentation regarding their prototypes. This was addressed by integrating information from a variety of reliable sources (two industry projects: Dr. Memory and CISCO Pyrebox), enabling the tool to support the correct tracing of more than 17,000 APIs. Moreover, a significant engineering effort had to be made in order to solve some of Intel Pin's inherent limitations which were encountered along the way.

Finally, the tests conducted on the Al-Khaser demonstration application confirmed BlueTracer's ability to provide extensive and informative logs while, at the same time, remaining undetected.

@@

A lot has to still be done in order to further improve BlueTracer's logging capabilities. In fact, the tool's ability to log the contents of complex data structures such as \texttt{structs} and \texttt{unions} is still limited and needs to be worked on. Also, the tracer does not deal with complex scenarios where there are dependencies between parameters, i.e., the type of a parameter depends on the value of another parameter. In addition, we would also be able to find a way to print the arguments values for user-mode callbacks and Windows Asychronous Procedure Calls by unpacking the data structures they receive as input. On top of that, BlueTracer, as of right now, supports only 32-bit Windows operating systems and, in light of this, we would like to extend its support for 64-bit Windows operating systems as well. It is worth noting that the tool was only tested with Windows 7, so it might be also a good a idea to evaluate it when running in other versions of the Windows operating system, even though this should not impact BlueTracer's functionalities.

Another future direction regarding BlueTracer might be the application of Big Data algorithms to the logs it produces. In fact, due to the wealth of information being recorded, the logs produced by BlueTracer are often very large (up to tens of gigabytes), making it challenging to extract valuable insights from them. In the specific case of BlueTracer, the information of interest is represented by the set of malicious and anti-analysis actions performed by malware. As a result of this, it might be a good idea to applying Big Data analytics to isolate just the relevant information, thus making the analyst's life easier.

>>>>

\chapter{Conclusions} \label{conclusion}

In order to deal with the astounding amount of malware instances surfacing every day, there has been in recent years a shift towards the use of dynamic malware analysis techniques over static ones. A popular dynamic analysis strategy is function call monitoring, as it enables the analyst to grasp the behavior of the malware sample under analysis, despite the presence of code obfuscations. Such technique is normally implemented via API hooking, that is, the interception of functions exported by dynamically linked libraries (DLL). However, API hooking has to face the growing trend of evasive malware strains, which conceal their damaging behavior when under the magnifying glass of an analyst. Furthermore, most of the current API hooking techniques presented in literature do not provide adequate countermeasures against evasive malware.

Motivated by this realization, in this thesis we have developed BlueTracer, a robust library and system call tracer for Windows applications specialized in evasive malware, which was implemented adopting the Intel Pin dynamic binary instrumentation framework and is capable of remaining hidden thanks to its integration with the BluePill stealthy execution toolkit. During the tool's development, a notable implementation challenge was posed by the heterogeneity of Windows libraries employed by malware and by the lack of appropriate documentation regarding their prototypes. This was addressed by integrating information from a variety of reliable sources (two industry projects: Dr. Memory and CISCO Pyrebox), enabling the tool to support the correct tracing of more than 17,000 APIs. Moreover, a significant engineering effort had to be made in order to solve some of Intel Pin's inherent limitations which we encountered along the way.

Finally, the tests conducted on the Al-Khaser anti-malware system stress tool confirmed BlueTracer's ability to provide extensive and informative logs while, at the same time, remaining undetected.

@@

A lot has to still be done in order to further improve BlueTracer's logging capabilities. In fact, the tool's ability to log the contents of complex data structures such as \texttt{structs} and \texttt{unions} is still limited and needs to be worked on. Also, the tracer does not deal with complex scenarios where there are dependencies between parameters, i.e., the type of a parameter depends on the value of another parameter. In addition, we should be able to find a way to print the arguments values for user-mode callbacks and Windows Asychronous Procedure Calls by unpacking the data structures they receive as input. On top of that, BlueTracer, as of right now, supports only 32-bit Windows operating systems and, in light of this, we would like to extend its support for 64-bit versions. It is worth noting that the tool was only tested with Windows 7, so it might be also a good a idea to evaluate it when running in other versions of the Windows operating system, even though this should not impact BlueTracer's functionalities.

Another future direction regarding BlueTracer might be the application of Big Data algorithms to the logs it produces. In fact, due to the wealth of information being recorded, the logs produced by BlueTracer are often very large (up to tens of gigabytes), making it challenging to extract valuable insights from them. In the specific case of BlueTracer, the information of interest is represented by the set of malicious and anti-analysis actions performed by malware. As a result of this, it might be a good idea to applying Big Data analytics to isolate just the relevant information and to even compare and integrate the data coming from different logs, thus making the analyst's life easier.

### 2018-07-09 evaluation.tex
Fixes Done

\begin{itemize}
\item \textbf{Empty Image.} All image notification functions are empty, that is, they just increase a global counter value. This means that no analysis routines are inserted at run-time. 
\item \textbf{Empty Routine.} All analysis routines are empty, i.e., all they do is, once again, increase a global counter value.
\item \textbf{Main Image.} Only Native APIs and APIs being directly invoked from the main executable of the pinned application are traced. This is BlueTracer's default mode of operation. 
\item \textbf{Complete.} Every event is traced, also including Native APIs and APIs being invoked outside the main executable (e.g., within libraries).
\end{itemize}

@@

The recorded execution times can be observed from \textit{Table 4.1}. As it can be seen, as more demanding modes of operation are employed, the run-time increases, although the increment is not uniform, but is actually quite variable across the different applications.

@@

\textit{Table 4.2} lists the overhead per event value of each benign application for which the execution time in the \textbf{Complete} mode of operation could be obtained. In this case we can see that the typical overhead is approximately in the 4$\times10^{-5}$ - 22$\times10^{-5}$ seconds range, with the compression programs being an exception, as they have a greater overhead per event value than the rest.

@@

To validate BlueTracer, Al-Khaser was run under it in the default \textbf{Main Image} mode. The result was that BlueTracer managed to remain undetected with respect to all the checks performed by Al-Khaser. From this outcome we obtained a first confirmation of BlueTracer's ability to hide its presence.

Furthermore, to also validate BlueTracer's tracing power, we analyzed the logs obtained when running Al-Khaser, with the objective of finding evidence of the performed evasion checks.
Even in this case, BlueTracer performed well, as the majority of the checks could easily be deduced from the log file.

@@

Anti-debugging checks are essentially aimed at determining if a program is running under a debugger. As already mentioned in Section~\ref{Debugger}, there exist Windows APIs whose objective is exactly to discover if the calling process is being debugged.
A well-known API of this kind is \texttt{IsDebuggerPresent}, which returns a nonzero value if the calling process is running in the context of a debugger and zero otherwise. Such API is employed in one of Al-Khaser's anti-debugging checks and is appropriately recorded by BlueTracer (\textit{Listing 4.1}). As it can be seen below, the return value is zero, as no debugger is detected.

@@

The \texttt{NtDelayExecution} Native API is lowest level user mode mechanism which can be employed to suspend execution. It takes as input two paramaters: a boolean that makes the Native API alertable when set to \texttt{true} and a \texttt{LARGE\_INTEGER} defining the delay interval. If the value for the second parameter is negative, it specifies the \textit{relative} amount to sleep in units of 100 nanoseconds; otherwise, it indicates an \textit{absolute time}, i.e., the date and time when to stop sleeping. Al-Khaser makes use of this Native API and BlueTracer was able to correctly trace it (\textit{Listing 4.4}). In particular, from the log entry, it can be seen that a negative delay was specified, meaning that a \textit{relative} time to sleep was provided.

@@

The mouse cursor position can be obtained through the use of the \texttt{GetCursorPos} API, which has a single output parameter containing a pointer to a \texttt{POINT struct}. Such data structure stores the retrieved position of the mouse cursor, in screen coordinates. Al-Khaser employs the \texttt{GetCursorPos} API to check if there is some kind of mouse movement in the sandbox. Even in this case, BlueTracer proved to work well as it traced the API's invocation (\textit{Listing 4.7}). It is interesting to notice that, from the log, it can be also deduced how the actual check is carried out by Al-Khaser. In fact, \texttt{GetCursorPos} is called, followed by a \texttt{Sleep} and another \texttt{GetCursorPos} invocation. In light of this, it is likely that an initial cursor position was first retrieved, which was later compared with the cursor position obtained after the sleep operation, with the aim of determining if the mouse cursor had moved in the mean time.

@@

A common way to detect virtualization is by checking Windows registry artifacts. The Windows registry is a hierarchical tree-like database containing critical data for both the Windows operating system and its applications. Each node of the tree is a \textit{key} and every key embodies \textit{subkeys} and \textit{values} \cite{Regs}. The existence of some specific registry keys is enough to reveal the presence of virtualization. For instance, the key \texttt{HKEY\_LOCAL\_MACHINE\textbackslash}\texttt{{HARDWARE}\textbackslash{ACPI}\textbackslash{DSDT}}\texttt{\textbackslash{VBOX}\textbackslash\_\_}
gives away the presence of VirtualBox. Al-Khaser looks for these kinds of Windows registry artifacts and BlueTracer is able to record such events (\textit{Listing 4.8}). From the listing below, it can be observed that the aforementioned registry key is searched through the use of the \texttt{RegOpenKeyExW} API, which normally opens the specified registry key. However, in this case, a file not found error value (\texttt{0x2}) is returned since the registry key is not found, thus also confirming BlueTracer's ability to remain hidden.

@@

A virtualized environment can be also discovered by looking for specific windows. The \texttt{FindWindow} API returns an handle of a window belonging to a specific class or having the specified name. This API can be exploited to check for the presence of a window associated to a virtualization program. A typical example of this trick, which is implemented by Al-Khaser, is searching for a window named "VBoxTrayToolWnd", as the presence of such window indicates that VirtualBox is being utilized. BlueTracer also detected this check (\textit{Listing 4.10}). As it can be seen, the return value is \texttt{null}, meaning that the window was not found and that BlueTracer did not reveal its presence.

@@

Lastly, anti-analysis checks are aimed at detecting the existence of processes belonging to popular analysis tools. One of such tools is Process Monitor (\texttt{procmon.exe}), which is an advanced Windows monitoring tool that shows real-time file system, registry and process/thread activity \cite{procmon}. Al-Khaser looks for the processes of many analysis tools, including the process associated to Process Monitor. Again, BlueTracer was also able to successfully log anti-analysis events, even the ones directed at uncovering the presence of Process Monitor (\textit{Listing 4.11}). From the listing below, it can be seen that the course of action adopted by Al-Khaser is quite straightforward. It first takes a snapshot of all the processes in the system with \texttt{CreateToolhelp32Snapshot}. Then, for each process in the snapshot, it retrieves the corresponding name, through the use of \texttt{Process32FirstW} at the beginning and \texttt{Process32NextW} afterwards, and compares it with \texttt{procmon.exe} via \texttt{StrCmpIW}.

@@

\section{Conclusions}
In this chapter we have outlined the results of BlueTracer's experimental evaluation. In order to assess BlueTracer's run-time overhead, we have tested it on a number of applications performing various tasks, employing a four different modes of operation. We have also introduced the overhead per event metric as a way of quantifying the run-time overhead introduced by BlueTracer. The results showed that the typical overhead per event is in the 4$\times10^{-5}$ - 22$\times10^{-5}$ seconds range.
Then we validated BlueTracer with Al-Khaser, an open-source application employed to assess how stealthy an analysis system is with respect to many evasion techniques used in the wild. BlueTracer was able to remain hidden to all the checks performed by Al-Khaser, thus proving to be effective at staying undetected. Furthermore, BlueTracer's tracking abilities also turned out to be efficacious, since evidence of the vast majority of Al-Khaser's techniques was present in BlueTracer's log.

>>>>

\begin{itemize}
\item \textbf{Empty Image.} All image notification functions are empty, that is, they just increase a global counter value. This means that no analysis routines are inserted at run-time. 
\item \textbf{Empty Routine.} All analysis routines are empty, i.e., all they do is, once again, increase a global counter value. So, here, with respect of \textbf{Empty Image}, there is an additional cost instrumentation cost derived from the insertion of analysis routines, which collect data but do not analyze it.  
\item \textbf{Main Image.} Only Native APIs and APIs being directly invoked from the main executable of the pinned application are traced. This is BlueTracer's default mode of operation. 
\item \textbf{Complete.} Every event is traced, also including Native APIs and APIs being invoked outside the main executable (e.g., within libraries).
\end{itemize}

@@

The recorded execution times can be observed in \textit{Table 4.1}. As the reader can see, as more demanding modes of operation are employed, the running time increases, although the increment is not uniform, but is actually quite variable across the different applications.

@@

\textit{Table 4.2} lists the overhead per event value of each benign application for which the execution time in the \textbf{Complete} mode of operation could be obtained. In this case we can see that the typical overhead is approximately in the range of 40 - 220 microseconds per operation, with the compression programs being an exception, as they have a greater overhead per event value than the rest.

@@

To validate BlueTracer, Al-Khaser was run under it in the default \textbf{Main Image} mode. The result was that BlueTracer managed to remain undetected thanks to its integration with BluePill with respect to all the checks performed by Al-Khaser. From this outcome we obtained a first confirmation of BlueTracer's ability to hide its presence.

Furthermore, to also validate BlueTracer's tracing power, we analyzed the logs obtained when running Al-Khaser, with the objective of finding evidence of the performed evasion checks.
Even in this case, BlueTracer performed well, as the checks could easily be deduced from the log file.

@@

Anti-debugging checks are essentially aimed at determining if a program is running under a debugger. As already mentioned in Section~\ref{Debugger}, there exist Windows APIs whose objective is exactly to discover if the calling process is being debugged.
A well-known API of this kind is \texttt{IsDebuggerPresent}, which returns a nonzero value if the calling process is running in the context of a debugger and zero otherwise. Such API is employed in one of Al-Khaser's anti-debugging checks and is properly recorded by BlueTracer (\textit{Listing 4.1}). As the reader can see below, the return value is zero, as no debugger is detected.

@@

The \texttt{NtDelayExecution} Native API is the lowest level user mode mechanism which can be employed to suspend execution. It takes as input two paramaters: a boolean that makes the Native API alertable when set to \texttt{true} and a \texttt{LARGE\_INTEGER} defining the delay interval. If the value for the second parameter is negative, it specifies the \textit{relative} amount to sleep in units of 100 nanoseconds; otherwise, it indicates an \textit{absolute time}, i.e., the date and time when to stop sleeping. Al-Khaser makes use of this Native API and BlueTracer was able to correctly trace it (\textit{Listing 4.4}). In particular, from the log entry, it can be seen that a negative delay was specified, meaning that a \textit{relative} time to sleep was provided.

@@

The mouse cursor position can be obtained through the use of the \texttt{GetCursorPos} API, which has a single output parameter containing a pointer to a \texttt{POINT struct}. Such data structure stores the retrieved position of the mouse cursor, in screen coordinates. Al-Khaser employs the \texttt{GetCursorPos} API to check if there is some kind of mouse movement in the sandbox. BlueTracer proved to work well as it traced the API's invocation (\textit{Listing 4.7}). It is interesting to notice that, from the log, it can be also deduced how the actual check is carried out by Al-Khaser. In fact, \texttt{GetCursorPos} is called, followed by a \texttt{Sleep} and another \texttt{GetCursorPos} invocation. In light of this, it is likely that an initial cursor position was first retrieved, which was later compared with the cursor position obtained after the sleep operation, with the aim of determining if the mouse cursor had moved in the mean time.

@@

A common way to detect virtualization is by checking Windows registry artifacts. The Windows registry is a hierarchical tree-like database containing critical data for both the Windows operating system and its applications. Each node of the tree is a \textit{key} and every key embodies \textit{subkeys} and \textit{values} \cite{Regs}. The existence of some specific registry keys is enough to reveal the presence of virtualization. For instance, the key \texttt{HKEY\_LOCAL\_MACHINE\textbackslash}\texttt{{HARDWARE}\textbackslash{ACPI}\textbackslash{DSDT}}\texttt{\textbackslash{VBOX}\textbackslash\_\_}
gives away the presence of VirtualBox. Al-Khaser looks for these kinds of Windows registry artifacts and BlueTracer is able to record such events (\textit{Listing 4.8}). From the listing below, it can be observed that the aforementioned registry key is searched through the use of the \texttt{RegOpenKeyExW} API, which normally opens the specified registry key. However, in this case, a "file not found" error value (\texttt{0x2}) is returned since the registry key is not found, thus also confirming BlueTracer's ability to remain hidden.

@@

A virtualized environment can be also discovered by looking for specific windows. The \texttt{FindWindow} API returns an handle of a window belonging to a specific class or having the specified name. This API can be exploited to check for the presence of a window associated to a virtualization program. A typical example of this trick, which is implemented by Al-Khaser, is searching for a window named "VBoxTrayToolWnd", as the presence of such window indicates that VirtualBox is being utilized. BlueTracer also detected this check (\textit{Listing 4.10}). As the reader can see, the return value is \texttt{null}, meaning that the window was not found and that BlueTracer did not reveal its presence.

@@

Lastly, anti-analysis checks are aimed at detecting the existence of processes belonging to popular analysis tools. One of such tools is Process Monitor (\texttt{procmon.exe}), which is an advanced Windows monitoring tool that shows real-time file system, registry and process/thread activity \cite{procmon}. Al-Khaser looks for the processes of many analysis tools, including Process Monitor. Again, BlueTracer was also able to successfully log anti-analysis events, even the ones directed at uncovering the presence of Process Monitor (\textit{Listing 4.11}). From the listing below, it can be seen that the course of action adopted by Al-Khaser is quite straightforward. It first takes a snapshot of all the processes in the system with \texttt{CreateToolhelp32Snapshot}. Then, for each process in the snapshot, it retrieves the corresponding name, through the use of \texttt{Process32FirstW} at the beginning and \texttt{Process32NextW} afterwards, and compares it with \texttt{procmon.exe} via \texttt{StrCmpIW}.

@@

\section{Conclusions}
In this chapter we have outlined the results of BlueTracer's experimental evaluation. In order to assess BlueTracer's run-time overhead, we have tested it on a number of applications performing various tasks, employing a four different modes of operation. We have also introduced the overhead per event metric as a way of quantifying the run-time overhead introduced by BlueTracer. The results showed that the typical overhead per event is in the range of 40 - 220 microseconds per operation.
Then we validated BlueTracer with Al-Khaser, an open-source application employed to assess how stealthy an analysis system is with respect to many evasion techniques used in the wild. BlueTracer was able to remain hidden to all the checks performed by Al-Khaser, thus proving to be effective at staying undetected. Furthermore, BlueTracer's tracking abilities also turned out to be efficacious, since evidence of the vast majority of Al-Khaser's techniques was present in BlueTracer's log.

### 2018-07-09 implementation.tex
Fixes Done

\subsection{Main Image Check} \label{MainImage}
In order to filter the logged information relative to Native APIs, BlueTracer allows the analyst to decide, through the use of a boolean parameter (\texttt{MainImage}) in the configuration file, whether only Native APIs invoked directly from the main executable of the pinned application should be traced. Therefore, as it can be seen in \textit{Figure~\ref{NTAPI Tracing}}, the first thing we do in \texttt{TraceSysCallEntry}, assuming that the \texttt{MainImage} configuration parameter has been set to \texttt{true}, is to determine if the Native API call is taking place directly from the main executable.

@@

APCs are of different types, one of which is user mode APCs. These usually call user mode code and are normally dispatched when the thread willingly enters into an alertable state. This means that, typically, user mode APCs do not asynchronously force themselves into the targeted thread, but they can be seen more like a set of work items in a queue, which the thread processes when he decides to do so \cite{APC1}.

>>>>

\subsection{Main Image Check} \label{MainImage}
In order to filter the logged information relative to Native APIs, BlueTracer allows the analyst to decide, through the use of a boolean parameter (\texttt{MainImage}) in the configuration file, whether only Native APIs invoked directly from the main executable of the pinned application should be traced. Therefore, as the reader can see in \textit{Figure~\ref{NTAPI Tracing}}, the first thing we do in \texttt{TraceSysCallEntry}, assuming that the \texttt{MainImage} configuration parameter has been set to \texttt{true}, is to determine if the Native API call is taking place directly from the main executable.

@@

APCs are of different types, one of which is user mode APCs. These usually call user mode code and are normally dispatched when the thread willingly enters into an alertable wait state. This means that, typically, user mode APCs do not asynchronously force themselves into the targeted thread, but they can be seen more like a set of work items in a queue, which the thread processes when he decides to do so \cite{APC1}.
